---
title: "Chapter 10 - Describing Discrete-Time Event Occurrence Data"
output: html_notebook
---

# The Life Table

The fundamental tool for summarizing the sample distribution of event occurrence is the life table. A life table tracks the event histories of a sample of individuals from the beginning of time through the end of data collection.

```{r}
library(tidyverse)

teachers.data <- read_csv("data/teachers.csv")

glimpse(teachers.data)
```

Build the life table:

```{r}
library(survival)

fit10.1 <- survfit(data = teachers.data, Surv(t, 1 - censor) ~ 1)

str(fit10.1)
```

The life table includes information on the number of individuals who:

* entered the interval
* experienced the target event during the interval
* were censored at the end of the interval

## A Framework for Characterizing the Distribution of Discrete-Time Event Occurrence Data

### Hazard Function

Hazard functions are used to assess the risk of event occurrence in each discrete time period. 

```{r}
most_rows <-
  tibble(year = fit10.1$time) %>% 
  mutate(time_int   = str_c("[", year, ", ", year + 1, ")"), 
         n_employed = fit10.1$n.risk, 
         n_left     = fit10.1$n.event) %>% 
  mutate(n_censored   = n_employed - n_left - lead(n_employed, default = 0),
         hazard_fun   = n_left / n_employed,
         survivor_fun = fit10.1$surv)

most_rows

row_1 <-
  tibble(year         = 0, 
         time_int     = "[0, 1)", 
         n_employed   = fit10.1$n.risk[1], 
         n_left       = NA, 
         n_censored   = NA, 
         hazard_fun   = NA, 
         survivor_fun = 1)

d <-
  bind_rows(row_1,
            most_rows)

d

d %>% 
  ggplot(aes(x = year, y = hazard_fun)) +
  geom_line() +
  scale_x_continuous("years in teaching", breaks = 0:13, limits = c(0, 13)) +
  scale_y_continuous(expression("estimated hazard probability, "*hat(italic(h))(italic(t[j]))), 
                     breaks = c(0, .05, .1, .15), limits = c(0, .15)) +
  theme(panel.grid = element_blank())
```

### Survivor Function

The survivor function cumulates the periord-by-period risks of event occurrence together to assess the probability that a randomly selected individual will "survive."

```{r}
d %>% 
  ggplot(aes(x = year, y = survivor_fun)) +
  geom_hline(yintercept = .5, color = "white", linetype = 2) +
  geom_line() +
  scale_x_continuous("years in teaching", breaks = 0:13, limits = c(0, 13)) +
  scale_y_continuous(expression("estimated survival probability, "*hat(italic(S))(italic(t[j]))),
                     breaks = c(0, .5, 1), limits = c(0, 1)) +
  theme(panel.grid = element_blank())
```

### Median Lifetime

# Developing Intuition About Hazard Functions, Survivor Functions, and Median Lifetimes

# A Simple and Useful Strategy for Constructing the Life Table

```{r}
teachers.data %>% 
  filter(id %in% c(20, 126, 129))

teachers.pp <-
  teachers.data %>% 
  uncount(weights = t) %>% 
  group_by(id) %>% 
  mutate(period = 1:n()) %>% 
  mutate(event = if_else(period == max(period) & censor == 0, 1, 0)) %>% 
  select(-censor) %>% 
  ungroup()

teachers.pp %>% 
  filter(id %in% c(20, 126, 129))

# count censored observations
teachers.pp %>%
  group_by(id) %>%
  summarise(period = max(period),
            event = last(event)) %>%
  count(event)

teachers.data %>%
  distinct(id) %>%
  count()
```

# Using the Person-Period Data Set to Construct the Life Table

```{r}
teachers_lt <-
  teachers.pp %>% 
  # change the coding for `event` in anticipation of the final format
  mutate(event = str_c("event = ", event)) %>% 
  group_by(period) %>% 
  count(event) %>% 
  ungroup() %>% 
  pivot_wider(names_from = event,
              values_from = n) %>% 
  mutate(total = `event = 0` + `event = 1`) %>% 
  mutate(prop_e_1 = (`event = 1` / total) %>% round(digits = 4))

teachers_lt
```

# Fit the Discrete-time Hazard Model with BRMS

```{r}
library(brms)
library(cmdstanr)

teachers_lt <- 
  teachers_lt %>%
  mutate(period = factor(period),
         event = `event = 1`)

fit10.6 <- 
  brm(event | trials(total) ~ 0 + period,
      family = binomial(link = "logit"),
      data = teachers_lt,
      prior(normal(0, 4), class = b),
      cores = 4,
      backend = "cmdstan")

print(fit10.6)

# convert coefficients back to the probability scale
fixef(fit10.6) %>% inv_logit_scaled()
```

Plot the hazard function:

```{r}
library(tidybayes)

posterior_samples(fit10.6) %>% 
  select(starts_with("b_")) %>% 
  mutate_all(inv_logit_scaled) %>% 
  set_names(1:12) %>% 
  pivot_longer(everything(),
               names_to = "period",
               values_to = "hazard") %>% 
  mutate(period = as.double(period)) %>% 
  
  ggplot(aes(x = period, y = hazard)) +
  stat_lineribbon(.width = c(.5, .95), size = 1/3) +
  # add the hazard estimates from `survival::survfit()`
  geom_point(data = tibble(period = fit10.1$time,
                           hazard = fit10.1$n.event / fit10.1$n.risk),
             aes(y = hazard),
             size = 2, color = "violetred1") +
  scale_fill_manual("CI", values = c("grey75", "grey60")) +
  scale_x_continuous(breaks = 1:12) +
  theme(legend.background = element_rect(fill = "transparent"),
        legend.key = element_rect(color = "grey92"),
        legend.position = c(.925, .825),
        panel.grid = element_blank())
```

Compare standarad errors:

```{r}
posterior_samples(fit10.6) %>% 
  select(starts_with("b_")) %>% 
  mutate_all(inv_logit_scaled) %>% 
  set_names(1:12) %>% 
  pivot_longer(everything(),
               names_to = "period",
               values_to = "hazard") %>% 
  mutate(period = period %>% as.double()) %>% 
  group_by(period) %>% 
  summarise(sd = sd(hazard)) %>% 
  bind_cols(se_h_hat %>% select(se_h_hat)) %>% 
  pivot_longer(-period) %>% 
  mutate(name = factor(name,
                       levels = c("sd", "se_h_hat"),
                       labels = c("Bayesian", "ML"))) %>% 
  
  ggplot(aes(x = period, y = value, color = name)) +
  geom_point(size = 3, position = position_dodge(width = .25)) +
  scale_color_viridis_d(NULL, option = "A", end = .55) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(expression(italic(se)), limits = c(0, 0.01)) +
  theme(legend.background = element_rect(fill = "transparent"),
        legend.key = element_rect(color = "grey92"),
        legend.position = c(.09, .9),
        panel.grid = element_blank())
```

