---
title: "Chapter 5 - Treating Time More Flexibly"
output: html_notebook
---

# Variably Spaced Measurement Occasions

## The Structure of Variably Spaced Data Sets

```{r}
library(tidyverse)

reading.pp <- read_csv("data/reading_pp.csv")
head(reading.pp)
```

Figure 5.1

```{r}
reading.pp %>%
  nest(data = c(wave, agegrp, age, piat)) %>%
  sample_n(size = 9) %>%
  unnest(data) %>%
  mutate(id = ifelse(id < 10, str_c("0", id), id) %>% str_c("id = ", .)) %>%
  pivot_longer(contains("age")) %>%
  ggplot(aes(x = value, y = piat, shape = name, linetype = name)) +
  geom_point(alpha = 2/3) +
  geom_smooth(method = "lm", se = F, size = 1/2) +
  facet_wrap(~id) +
  scale_shape_manual(values = c(19, 3)) +
  xlab("measure of age") +
  coord_cartesian(xlim = c(5, 12),
                  ylim = c(0, 80)) +
  theme(panel.grid = element_blank())
```

## Postulating and Fitting Multilevel Models with Variably Spaced Waves of Data

```{r}
reading.pp <- 
  reading.pp %>%
  mutate(agegrp_c = agegrp - 6.5,
         age_c = age - 6.5)

head(reading.pp)

library(brms)
library(cmdstanr)

fit5.1 <- 
  brm(piat ~ 0 + Intercept + agegrp_c + (1 + agegrp_c | id),
      family = gaussian,
      prior = c(prior(normal(100, 30), class = b, coef = Intercept),
                prior(normal(0, 30), class = b, coef = agegrp_c),
                prior(student_t(3, 0, 15), class = sd),
                prior(student_t(3, 0, 15), class = sigma),
                prior(lkj(4), class = cor)),
      data = reading.pp,
      cores = 4,
      backend = "cmdstan")
print(fit5.1, digits = 3)

fit5.2 <- 
  brm(piat ~ 0 + Intercept + age_c + (1 + age_c | id),
      family = gaussian,
      prior = c(prior(normal(100, 30), class = b, coef = Intercept),
                prior(normal(0, 30), class = b, coef = age_c),
                prior(student_t(3, 0, 15), class = sd),
                prior(student_t(3, 0, 15), class = sigma),
                prior(lkj(4), class = cor)),
      data = reading.pp,
      cores = 4,
      backend = "cmdstan")
```

# Varying Number of Measurement Occasions

## Analyzing Data Sets in Which the Number of Waves per Person Varies

```{r}
wages.pp <- read_csv("data/wages_pp.csv")

glimpse(wages.pp)

wages.pp %>% 
  select(id, exper, lnw, black, hgc, uerate) %>% 
  filter(id %in% c(206, 332, 1028))
```

Here's how we can look at the number of occasions per `id`.

```{r}
wages.pp %>%
  group_by(id) %>%
  count() %>%
  ggplot(aes(x = n)) +
  geom_bar() +
  scale_x_continuous("# measurement occasions", breaks = 1:13) +
  ylab("count of cases") +
  coord_flip() +
  ggpubr::theme_pubr()

wages.pp %>%
  filter(id %in% c(206, 332, 1028)) %>%
  mutate(id = factor(id)) %>%
  ggplot(aes(x = exper, y = lnw, color = id)) +
  geom_point() +
  geom_line() +
  ggpubr::theme_pubr()
```

Fitting the model.

```{r}
fit5.3 <-
  brm(data = wages.pp, 
      family = gaussian,
      lnw ~ 0 + Intercept + exper + (1 + exper | id),
      prior = c(prior(normal(1.335, 1), class = b, coef = Intercept),
                prior(normal(0, 0.5),   class = b, coef = exper),
                prior(student_t(3, 0, 1), class = sd),
                prior(student_t(3, 0, 1), class = sigma),
                prior(lkj(4), class = cor)),
      cores = 4,
      backend = "cmdstan")
print(fit5.3, digits = 3)

post <-
  posterior_samples(fit5.3) %>% 
  transmute(percent_change = (exp(b_exper) - 1))

head(post)
```

Plot the model's results.

```{r}
library(tidybayes)

post %>%
  ggplot(aes(x = percent_change, y = 0)) +
  geom_halfeyeh(.width = 0.95) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Percent Change",
       x = expression(100*(italic(e)^(hat(gamma)[1][0])-1))) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggpubr::theme_pubr()
```

Now we add two level-2 covariates: `black` and `hgc.9`.

```{r}
wages.pp <- 
  wages.pp %>%
  rename(hgc_9 = hgc.9)

wages.pp %>%
  pivot_longer(c(black, hgc_9)) %>%
  ggplot(aes(x = value)) +
  geom_bar() +
  facet_wrap(~ name, scales = "free") +
  ggpubr::theme_pubr() 
```

```{r}
fit5.4 <-
  brm(data = wages.pp, 
      family = gaussian,
      lnw ~ 0 + Intercept + hgc_9 + black + exper + exper:hgc_9 + 
        exper:black + (1 + exper | id),
      prior = c(prior(normal(1.335, 1), class = b, coef = Intercept),
                prior(normal(0, 0.5),   class = b),
                prior(student_t(3, 0, 1), class = sd),
                prior(student_t(3, 0, 1), class = sigma),
                prior(lkj(4), class = cor)),
      cores = 4,
      backend = "cmdstan")
print(fit5.4, digits = 3)

post <- posterior_samples(fit5.4) 

post %>% 
  transmute(`sigma[0]^2` = sd_id__Intercept^2,
            `sigma[1]^2` = sd_id__exper^2,
            `sigma[epsilon]^2` = sigma^2) %>% 
  pivot_longer(everything()) %>% 
  
  ggplot(aes(x = value, y = name)) +
  geom_halfeyeh(.width = .95, normalize = "xy") +
  scale_y_discrete(NULL, labels = ggplot2:::parse_safe) +
  coord_cartesian(ylim = c(1.4, 3.4)) +
  theme(axis.ticks.y = element_blank(),
        panel.grid = element_blank())
```

```{r}
fit5.5 <-
  brm(data = wages.pp, 
      family = gaussian,
      lnw ~ 0 + Intercept + hgc_9 + exper + exper:black + (1 + exper | id),
      prior = c(prior(normal(1.335, 1), class = b, coef = Intercept),
                prior(normal(0, 0.5),   class = b),
                prior(student_t(3, 0, 1), class = sd),
                prior(student_t(3, 0, 1), class = sigma),
                prior(lkj(4), class = cor)),
      cores = 4,
      backend = "cmdstan")

nd <- 
  expand_grid(black = 0:1,
              hgc_9 = c(0, 3),
              exper = seq(from = 0, to = 11, length.out = 30))

f <- 
  fitted(fit5.5,
         newdata = nd,
         re_formula = NA) %>%
  data.frame() %>%
  bind_cols(nd)

head(f)
```

Figure 5.2

```{r}
f %>%
  mutate(black = factor(black,
                        labels = c("Latinos and Whites", "Blacks")),
         hgc_9 = factor(hgc_9, 
                        labels = c("9th grade dropouts", "12th grade dropouts"))) %>% 
  
  ggplot(aes(x = exper,
             color = black, fill = black)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5),
              size = 0, alpha = 1/4) +
  geom_line(aes(y = Estimate)) +
  scale_fill_viridis_d(NULL, option = "C", begin = .25, end = .75) +
  scale_color_viridis_d(NULL, option = "C", begin = .25, end = .75) +
  ylab("lnw") +
  coord_cartesian(ylim = c(1.6, 2.4)) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~hgc_9)
```

## Practical Problems That May Arise When Analyzing Unbalanced Data Sets

### Boundary Constraints

```{r}
wages.small.pp <- read_csv("data/wages_small_pp.csv") %>%
  rename(hgc_9 = hcg.9)

head(wages.small.pp)

wages.small.pp %>% 
  group_by(id) %>% 
  count() %>% 
  ggplot(aes(x = n)) +
  geom_bar() +
  scale_x_continuous("# measurement occasions", breaks = 1:13, limits = c(.5, 13)) +
  ylab("count of cases") +
  coord_flip() +
  theme(panel.grid = element_blank())
```

```{r}
fit5.6 <-
  brm(data = wages.small.pp, 
      family = gaussian,
      lnw ~ 0 + Intercept + hgc_9 + exper + exper:black + (1 + exper | id),
      prior = c(prior(normal(1.335, 1), class = b, coef = Intercept),
                prior(normal(0, 0.5),   class = b),
                prior(student_t(3, 0, 1), class = sd),
                prior(student_t(3, 0, 1), class = sigma),
                prior(lkj(4), class = cor)),
      cores = 4,
      backend = "cmdstan")
print(fit5.6, digits = 3)

fit5.7 <-
  brm(data = wages.small.pp, 
      family = gaussian,
      lnw ~ 0 + Intercept + hgc_9 + exper + exper:black + (1 + exper | id),
      prior = c(prior(normal(1.335, 1), class = b, coef = Intercept),
                prior(normal(0, 0.5),   class = b),
                prior(student_t(3, 0, 1), class = sd),
                prior(student_t(3, 0, 1), class = sigma),
                prior(lkj(4), class = cor)),
      cores = 4,
      control = list(adapt_delta = 0.99),
      backend = "cmdstan")
```

# Time-Varying Predictors

## Including the Main Effect of a Time-Varying Predictor

```{r}
unemp.pp <- read_csv("data/unemployment_pp.csv")

head(unemp.pp)

unemp.pp %>%
  distinct(id) %>%
  count()

# compute the number of participants who were never employed during the study
unemp.pp %>%
  filter(unemp == 0) %>%
  distinct(id) %>%
  count() %>%
  summarise(never_employed = 254 - n)

unemp.pp %>%
  ggplot(aes(x = months)) +
  geom_vline(xintercept = c(3, 8), color = "black") +
  geom_histogram(binwidth = 0.5) +
  ggpubr::theme_pubr()

# create an interview variable
unemp.pp <- 
  unemp.pp %>% 
  mutate(interview = ifelse(months < 3, 1, 
                            ifelse(months > 8, 3, 2)),
         interview_alt = case_when(months < 3 ~ 1,
                                   months > 8 ~ 3,
                                   months <= 8 & months >= 3 ~ 2))

unemp.pp %>%
  ggplot(aes(x = interview_alt)) +
  geom_bar()
```

Fit the unconditional growth model.

```{r}
fit5.11 <-
  brm(data = unemp.pp, 
      family = gaussian,
      cesd ~ 0 + Intercept + months + (1 + months | id),
      prior = c(prior(normal(14.5, 20), class = b, coef = Intercept),
                prior(normal(0, 10), class = b),
                prior(student_t(3, 0, 10), class = sd),
                prior(student_t(3, 0, 10), class = sigma),
                prior(lkj(4), class = cor)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      backend = "cmdstan",
      control = list(adapt_delta = .99))
print(fit5.11, digits = 3)

posterior_samples(fit5.11) %>%
  transmute(`first day of job loss` = b_Intercept,
            `linear decline by month` = b_months) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value, y = 0)) +
  geom_halfeyeh(.width = 0.95, normalize = "panels") +
  facet_wrap(~ name, scales = "free") +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab("CES-D Composite Score") +
  ggpubr::theme_pubr()

plot(conditional_effects(fit5.11), plot = FALSE)[[1]] +
  geom_hline(yintercept = 14.5, color = "grey50", linetype = 2) +
  coord_cartesian(ylim = c(0, 20)) +
  ggpubr::theme_pubr()
```

### Using a Composite Specification

```{r}
fit5.12 <-
  brm(data = unemp.pp, 
      family = gaussian,
      cesd ~ 0 + Intercept + months + unemp + (1 + months | id),
      prior = c(prior(normal(14.5, 20), class = b, coef = Intercept),
                prior(normal(0, 10), class = b),
                prior(student_t(3, 0, 10), class = sd),
                prior(student_t(3, 0, 10), class = sigma),
                prior(lkj(4), class = cor)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      backend = "cmdstan",
      control = list(adapt_delta = .95))
print(fit5.12)
```

Upper left panel of figure 5.3.

```{r}
nd <- 
  tibble(unemp = 1, 
         months = seq(from = 0, to = 14, by = 0.5))

f <- 
  fitted(fit5.12, newdata = nd, re_formula = NA) %>%
  data.frame() %>%
  bind_cols(nd)

ggplot(data = f, aes(x = months)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "grey67", alpha = 1/2) +
  geom_line(aes(y = Estimate)) +
  scale_x_continuous("Months since job loss", 
                     breaks = seq(from = 0, to = 14, by = 2)) +
  scale_y_continuous("CES-D", limits = c(5, 20)) +
  labs(subtitle = "Remain unemployed") +
  ggpubr::theme_pubr()
```

Upper right panel of figure 5.3.

```{r}
nd <- 
  tibble(unemp = rep(1:0, times = c(11, 19)),
         months = c(seq(from = 0, to = 5, by = 0.5),
                    seq(from = 5, to = 14, by = 0.5)))

f <- 
  fitted(fit5.12, newdata = nd, re_formula = NA) %>%
  data.frame() %>%
  bind_cols(nd)

ggplot(data = f, aes(x = months, group = unemp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "grey67", alpha = 1/2) +
  geom_line(aes(y = Estimate)) +
  geom_segment(x = 5, xend = 5,
               y = fixef(fit5.12)[1, 1] + fixef(fit5.12)[2, 1] * 5,
               yend = fixef(fit5.12)[1, 1] + fixef(fit5.12)[2, 1] * 5 + fixef(fit5.12)[3, 1],
               size = 1/3, linetype = 2) +
  annotate(geom = "text", x = 8, y = 14.5, label = "gamma[2][0]",
           parse = TRUE) +
  geom_segment(x = 7, xend = 5.5, y = 14.5, yend = 14.5, 
               arrow = arrow(length = unit(0.05, "inches"))) +
  scale_x_continuous("Months since job loss", breaks = seq(from = 0, to = 14, by = 2)) +
  scale_y_continuous("CES-D", limits = c(5, 20)) +
  labs(subtitle = "Reemployed at 5 months") +
  ggpubr::theme_pubr()
```
